Efficient Thresholded Correlation using Truncated Singular Value Decomposition

### 2015-11-20 threshold/draft.Rnw
Paper update
Minor re-wording of text, fix typos, etc. I am now thinking of avoiding discussion of probabalistic implications of reducint ell, in favor of a simple, crisp and concise result.

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasable for high-dimensional data arising in modern genomics and other
applications. Considerable attention has been devoted to methods called
\emph{pruning methods} that cheaply prune away all but the most
highly-correlated pairs of vectors.  A variety of pruning methods have been
described for the thresholded correlation problem.  For example the TAPER and
related algorithms of Xiong et al. base pruning rules on the frequency of
mutually co-ocurring observations (rows) among vector pairs; see \cite{prune1,
prune2} and the references therein. Related methods only approximate a set
containing the most highly-correlated pairs using hashing
algorithms\cite{prune3}. A number of other methods arrive at approximate sets
of the most highly-correlated pairs using dimensionality-reduction techniques
like the discrete Fourier transform\cite{timeseries}. The method of Wu, et
al.\cite{svd-similarity}, conceptually very similar to our approach, uses the
singular value decomposition (SVD) to find all pairs of close vectors with
respect to a distance metric (instead of correlation).  Our method finds all of
the most highly-correlated vector pairs with respect to a given threshold by
pruning along an intuitive path of decreasing variance defined by the data and
computed by the SVD.

@@

Let $A=U\Sigma V^T$ be the singular value decomposition of $A$, where
$U\in\mathcal{R}^{m\times k}$, $V\in\mathcal{R}^{n\times k}$,
$U^TU = V^TV = I$,
and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
later convenience, we write $s_{1:p}=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$
to be the vector of the first $p\le k$ singular values along the diagonal of
$\Sigma$.  Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
$v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
V\Sigma^2 V^T$ is a symmetric eigenvalue decomposition of the correlation
matrix with nonzero eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$.
The span of the columns of $V$ form an orthonormal basis of the range of the
correlation matrix $A^TA$.

@@

The following lemma expresses the Euclidean distance between any two columns of
the matrix $A$ as a weighted sum of projected interval distances along the SVD
basis vectors forming the columns of $V$.
\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{vdist}
\|a_i - a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} - e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{k} - e_j^Tv_{k})^2,
\end{equation}
\end{lemma}
where here and below, $e_{j}\in\mathcal{R}^n$ represents the $j$th unit basis
vector consisting of $1$ in position $j$ and zeros otherwise.
(Thus, $e_j^Tv_{k}$ is the $j$th position in the vector $v_k$; that is, the
$[j,k]$ entry in the matrix $V$.)

@@

Note that each term in the sum is nonnegative and the weights $\sigma_j^2$,
$j=1,2,\ldots,k$ are nonincreasing and defined by the covariance structure of
the data.  Given a correlation threshold $t$, $0<t<1$, Equations \ref{cordist}
and \ref{vdist} together suggest that pairs of vectors can be excluded from
consideration whenever their projected distance is too large.  For example if
$\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 > 2(1-t)$, then we can conclude from
just two scalar values that the column vectors $a_i$ and $a_j$ do not meet the
correlation threshold $t$.
In practice, many pairs of vectors may be pruned in this fashion by
inspecting distances in low-dimensional subspaces, substantially reducing the
complexity of computing thresholded correlation matrices.

@@

Let $t$ be a given correlation threshold $0<t<1$, and let
$P\in\mathcal{R}^{n\times n}$ be a permutation matrix that orders then entries
of $v_1$ in increasing order.  It follows from Equations \ref{cordist} and
\ref{vdist} that for any $i,j=1,2,\ldots,n$, if $\sigma_1^2 (e_i^Tv_1 -
e_j^Tv_1)^2 > 2(1-t)$ then $\mathrm{cor}(a_i,a_j) < t$.  Let $\ell$ be the
longest run of successive points in $P v_1$ within the interval
$\sqrt{2(1-t)}/\sigma_1$. For example, Figure \ref{fig1} displays the entries
of $v_1$ for the small ``BicatYeast'' example (where $A\in\mathcal{R}^{70\times
419}$) from Section \ref{examples} in increasing order.  The lines in the plot
illustrate the interval associated with the correlation threshold $t=0.95$. The
quantity $\ell$ is obtained by rolling the interval across all the points in
the plot and counting the maximum number of points that lie within the
interval.  With respect to the ordering $P$, $\ell$ represents the biggest
index difference that pairs of indices corresponding to correlated vectors
above the threshold can exhibit.

@@

@@

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$ with columns scaled to have
zero mean and unit norm, correlation threshold $t$,
truncated SVD rank $p<<k$.
\begin{enumerate}
\item Compute the truncated SVD
$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where\\
$V_{1:p}=[v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\},
\]
where $\tilde{j}=1,2,\ldots,n-\tilde{i}$.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$ and
restart IRLBA to compute 
$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$ and go to step 4.
Continue to increase $p$ in this way until the number of candidate
pairs is sufficiently small or stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

@@

We have so far assumed that the mean of each column vector of the $m\times n$
matrix $A$ is zero and the columns are scaled to have unit norm, but we really
want a method that works with general matrices $A$.  We also want a way to
efficiently compute a relatively low-rank truncated SVD of potentially large
matrices required by step 1 of Algorithm \ref{prune}.  Fortunately, both
desires are met by one algorithm.

@@

We can relax the column mean and scale assumptions without introducing much
additional computational or storage overhead as follows.  Let
$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
of the $m\times n$ matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
a diagonal matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLB to correlation
problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
at most computing two additional vector inner products per matrix-vector product.

@@

@@

\section{NOTES}

>>>>

The number of correlation coefficients grows quadratically with the number of
vectors. Simply computing all pairs of correlation coefficients and then
filtering out coefficients below a given threshold may not be computationally
feasable for high-dimensional data in modern genomics and other applications.
Considerable attention has been devoted to methods called \emph{pruning
methods} that cheaply prune away all but the most highly-correlated pairs of
vectors.  For example the TAPER and related algorithms of Xiong et al. base
pruning rules on the frequency of mutually co-ocurring observations (rows)
among vector pairs; see \cite{prune1, prune2} and the references therein.
Related methods only approximate a set containing the most highly-correlated
pairs using hashing algorithms\cite{prune3}. A number of other methods arrive
at approximate sets of the most highly-correlated pairs using
dimensionality-reduction techniques like the discrete Fourier
transform\cite{timeseries}. The method of Wu, et al.\cite{svd-similarity},
conceptually very similar to our approach, uses the singular value
decomposition (SVD) to find all pairs of close vectors with respect to a
distance metric (instead of correlation).  Our method finds all of the most
highly-correlated vector pairs with respect to a given threshold by pruning
along an intuitive path of decreasing variance defined by the data and computed
by the SVD.

@@

Let $A=U\Sigma V^T$ be a singular value decomposition of $A$, where
$U\in\mathcal{R}^{m\times k}$, $V\in\mathcal{R}^{n\times k}$,
$U^TU = V^TV = I$,
and $\Sigma\in\mathcal{R}^{k\times k}$ is a diagonal matrix with
diagonal entries $\sigma_1 \ge \sigma_2 \ge \cdots \ge \sigma_k > 0$.  For
later convenience, we write $s_{1:p}=[\sigma_1, \sigma_2, \ldots, \sigma_p]^T$
to be the vector of the first $p\le k$ singular values along the diagonal of
$\Sigma$.  Denote the columns of $V$ as $V=[v_1, v_2, \ldots, v_k]$, where each
$v_j\in\mathcal{R}^n$, $j=1,2,\ldots,k$.  Then $\mathrm{cor}(A) = A^TA =
V\Sigma^2 V^T$ is a symmetric eigenvalue decomposition of the correlation
matrix with nonzero eigenvalues $\sigma_1^2, \sigma_2^2, \ldots, \sigma_k^2$.
The span of the columns of $V$ form an orthonormal basis of the range of the
correlation matrix $A^TA$.

@@

The following lemma expresses the Euclidean distance between any two columns of
the matrix $A$ as a weighted sum of projected interval distances along the SVD
basis vectors forming the columns of $V$.
\begin{lemma}
Let $a_i$ and $a_j$, $1\le i,j\le n$ be any two columns of the matrix
$A$. Then
\begin{equation}\label{vdist}
\|a_i - a_j\|^2 =
\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 + 
\sigma_2^2 (e_i^Tv_{2} - e_j^Tv_{2})^2 + \cdots + 
\sigma_k^2 (e_i^Tv_{k} - e_j^Tv_{k})^2,
\end{equation}
\end{lemma}
where here and below, $e_{j}\in\mathcal{R}^n$ represents the $j$th unit basis
vector consisting of $1$ in position $j$ and zeros otherwise.
(Thus, $e_j^Tv_{k}$ is the $j$th position in the vector $v_k$--that is, the
$[j,k]$ entry in the matrix $V$.)

@@

Note that each term in the sum is nonnegative and the weights $\sigma_j^2$,
$j=1,2,\ldots,k$ are nonincreasing and defined by the covariance structure of
the data.  Given a correlation threshold $t$, $0<t<1$, Equations \ref{cordist}
and \ref{vdist} together suggest that pairs of vectors can be excluded from
consideration whenever their projected distance is too large.  For example if
$\sigma_1^2 (e_i^Tv_{1} - e_j^Tv_{1})^2 > 2(1-t)$, then we can conclude from
just a few scalar values that the column vectors $a_i$ and $a_j$ do not meet the
correlation threshold $t$.
In practice, many pairs of vectors may be pruned in this fashion by
inspecting distances in low-dimensional subspaces, substantially reducing the
complexity of computing thresholded correlation matrices.

@@

Let $t$ be a given correlation threshold, $0<t<1$, and let
$P\in\mathcal{R}^{n\times n}$ be a permutation matrix that orders the entries
of $v_1$ in increasing order.  For example, Figure \ref{fig1} displays the
ordered entries of $Pv_1$ for the small ``BicatYeast'' example (where
$A\in\mathcal{R}^{70\times 419}$) from Section \ref{examples}.  The lines in
the plot illustrate the interval associated with the correlation threshold
$t=0.95$ placed at an arbitrary vertical axis location.

@@

Let $\ell$ be the longest run of successive points in $P v_1$ within the
interval $\sqrt{2(1-t)}/\sigma_1$.  The quantity $\ell$ may, for example, be
obtained by rolling the interval over all the points in Figure \ref{fig1} and
counting the maximum number of points that lie within the interval.  With
respect to the ordering $P$, $\ell$ represents the biggest index difference
that pairs of indices corresponding to correlated vectors above the threshold
can exhibit.

@@

\begin{algorithm}\label{prune}
Input: data matrix $A\in \mathcal{R}^{m\times n}$ with columns scaled to have
zero mean and unit norm, correlation threshold $t$,
truncated SVD rank $p<<k$.
\begin{enumerate}
\item Compute the truncated SVD
$AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$,
where\\
$V_{1:p}=[v_1, v_2, \ldots, v_p]$
and $\mathrm{diag}(s_{1:p})$ denotes the diagonal matrix of the first $p$ singular values.
\item Compute permutation $P$ that orders points in $v_1$.
\item Compute $\ell$, the longest run of successive points in $P v_1$ within the interval $\sqrt{2(1-t)}/\sigma_1$.
\item Compute a set of candidate index pairs $(\tilde{j},\tilde{j}+\tilde{i})$ with respect to the index order defined by the permutation $P$ that possibly meet the correlation threshold
\[
\bigcup_{\tilde{i}=1}^\ell
\left\{
(\tilde{j},\tilde{j}+\tilde{i}) : 
e_{\tilde{j}}^T(D^{\tilde{i}} P V_{1:p} )^2 s_{1:p}^2 \le 2(1-t)
\right\},
\]
where $\tilde{j}=1,2,\ldots,n-\tilde{i}$.
\item If there are judged to be  ``too many'' candidate pairs, increase $p$,
compute $AV_{1:p} = U_{1:p}\mathrm{diag}(s_{1:p})$ and go to step 4.
Continue to increase $p$ in this way until the number of candidate
pairs is sufficiently small or stops decreasing (the best this algorithm can do).
\item Recover the non-permuted candidate column vector
indices $(j,i)$ from $j=e_{\tilde{j}}^T Pq$ and $i=e_{\tilde{j}+\tilde{i}}^TPq$ for every
pair $(\tilde{j},\tilde{j} + \tilde{i})$ in step 4, where
$q\in\mathcal{R}^n = [1,2,\ldots,n]^T$.
\item Compute the full correlation coefficient for each column vector pair identified in
the previous step, applying the correlation threshold to this reduced set of values.
\end{enumerate}
\end{algorithm}
Because the singular values $\sigma_j$, $j=1,2,\ldots,n$ are non-increasing and
decrease in proportion to variability in the data, it's often reasonable to
judge whether there are ``too many'' candidate pairs (step 5) after only a 1-d
projection $\tilde{i}=1$ in step 4 of the algorithm.

@@

We have so far assumed that the mean of each column vector of the $m\times n$
matrix $A$ is zero and the columns are scaled to have unit norm, but we really
want a method that works with general matrices $A$.  We also want a way to
efficiently compute a relatively low-rank truncated SVD of potentially large
matrices required by step 1 of Algorithm \ref{prune}.  And the truncated SVD
method should be able to cheaply restart to compute additional singular vectors
as required in step 5 of Algorithm \ref{prune}. Fortunately all desires are met
by one algorithm.

@@

We can relax the column mean and scale assumptions without introducing much
additional computational or storage overhead as follows. Assume $A$ is an
$m\times n$ real-valued matrix without constant-valued columns. Let
$z=[z_1,z_2,\ldots,z_n]\in\mathcal{R}^n$ represent the vector of column means
of the $m\times n$ matrix $A$, $w=[w_1^2,w_2^2,\ldots,w_n^2]\in\mathcal{R}^n$
be the vector of squared column norms of $A$, and $W\in\mathcal{R}^{n\times n}$
a diagonal matrix with diagonal entries $1/\sqrt{w_i^2 - m z_i^2}$.  Let
$e\in\mathcal{R}^m$ be a vector of all ones.  Then $(A - ez^T)W$ is a centered
matrix with zero column means and scaled to have unit column norms, and
\[
\mathrm{cor}(A) = W^T (A-ez^T)^T (A-ez^T) W.
\]
The IRLBA, based on the Lanczos process, is a method of iterated matrix-vector
products. The main idea behind efficient application of IRLB to correlation
problems replaces matrix-vector products of the form $Ax$ with $AWx - ez^TWx$ in
the iterations, implicitly working with a scaled and centered matrix without
forming it.  This comes at the cost of storing two additional length $n$ vectors and
at most computing two additional vector inner products per matrix-vector product.
The R implementation\cite{irlbar} of IRLBA includes arguments for centering and
scaling the input matrix.

@@

A full/naive thresholded correlation computation requires $O(n^2 m)$ flops. Let
$p$ be the selected IRLBA dimension. Then the proposed method requires
$O(\ell n p)$ flops, where $\ell$ is the longest run of ordered entries in $v_1$
that
meet the 1-d projected distance threshold $\sqrt{2(1-t)}/\sigma_1$,
plus the flops required by IRLBA and the post-processing step of evaluating the
correlation of all the candidate pairs.

@@

\section{NOTES not for publication}
